\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-06-05 coqimpl.tex
Save.

Our emphasis here is that, we singled out some programming language features from STLC and prove the type safety for each feature separately, following the examples of MTC. And we use a \textit{not-yet-completely-defined} \textbf{mix-in} feature to mix the semantic and properties of two programming language feature -- product and boolean -- with vanilla STLC. This way, we can say precisely, \textit{a programming language feature itself is a piece of data/inheritance judgement/inherited family}. Compared to MTC, our example uses small-step operational semantic by exploiting the extensible inductive type, and most of the proof is directly adapted from the one in Software Foundation, resulting a more accessible proof.

@@

\subsection{Abstract Interpreters for Imp}
The second example is adapted and modified from the Familia paper -- contrary to our first example, we use big-step interpreter and fuel to indicate the operational semantic on an imperative language with side-effect, and we specify the abstract interpretation and prove its soundness, with some of postulation on both computation and property. Then we extend the language feature and we instantiate the postulation on computation for both concrete interpreters and abstract interpreters. Thanks to the compilation, we can directly run the resulting abstract interpreters.

>>>>

Our emphasis here is that, we singled out some programming language features from STLC and prove the type safety for each feature separately, following the examples of MTC\citep{delaware2013,forsta2020}. And we use a \textit{not-yet-completely-defined} \textbf{mix-in} feature to mix the semantic and properties of two programming language feature -- product and boolean -- with vanilla STLC. This way, we can say precisely, \textit{a programming language feature itself is a piece of data/inheritance judgement/inherited family}. Compared to MTC, our example uses small-step operational semantic by exploiting the extensible inductive type, and most of the proof is directly adapted from the one in Software Foundation, resulting a more accessible proof. 
% In related work, Coq/Metatheory a la carte/Tion embedding can be emphasized as a more "semantic approach" because they encode the meaning using a special design pattern (for example, open recursive inductive type for extension), compared to our more syntactic approach. Their advantage is the transferability of this technique accross different proof assistants, and their disadvantage is that their approach are less accessible and unfriendly to amateur Coq users -- which can be reflected from the distance of their approach and text-book Software Foundation proof.

@@

\subsection{Abstract Interpreters for Imp}
The second example is adapted and modified from the Familia paper
% citation needed here
-- contrary to our first example, we use big-step interpreter and fuel to indicate the operational semantic on an imperative language with side-effect, and we specify the abstract interpretation and prove its soundness, with some of postulation on both computation and property. Then we extend the language feature and we instantiate the postulation on computation for both concrete interpreters and abstract interpreters. Thanks to the compilation, we can directly run the resulting abstract interpreters.

### 2022-06-05 standardmodel.tex
Save.

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from \citep{kaposi2017type}, where they formally resort to the concept of \textit{algebra} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities for accessibility\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.} The main idea of the proof is that: 1. we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; 2. we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-defined-ness).

@@

We omit the model here for extensible inductive type and inheritance judgement because they require inductive facility in meta-logic. In set theory, that would be using \textit{rule sets}\citep{timany2017consistency,aczel1998relating}, which would be more confusing for the reader. In appendix we will use the native inductive notion from Agda to construct models.

But we can sketch out their construction -- 1. for extensible W type, we simply construct a similar inductive type reflect the same idea, and the challenge mainly lies in satisfying the substitution law; 2. for inheritance judgement, we inductively define an almost-identical syntactic structure as model, but not a QIIT, and we inductively define meta-level substitution to implement the explicit substitution. We also define $inh$ operation by induction strictly following $\beta$-law.

Notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot \Rightarrow \denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.

>>>>

Once we have the syntax of our theory, the first question is if our theory is consistent, i.e., if we can syntactically derive bottom in our theory. We prove the consistency by extending the standard model from \citep{kaposi2017type, kaposi2019gluing}, where they formally resort to the concept of \textit{algebra} of QIIT, here we only consider inductively interpreting each syntax piece into naive set theory that is also respecting judgemental equalities for accessibility\footnote{In our appendix, we still interpret into Agda component to be more precise and less error-prone, also we need the inductive facility from Agda.} The main idea of the proof is that: 1. we use $\denotesS{}$ to map each intrinsic well-typed syntax piece into a mathematical object; 2. we define proof-irrelevant proposition $\models_S$ for each syntactical $\vdash$, and derive compatibility lemma. We will mainly focus on the definition of $\denotesS{}$ and omit the verification of $\models_S$\footnote{Actually our 1 is doing exactly an interpretation function, and in our agda formulation we only do 1 because 2 is directly implied by 1; we make 2 explicit here to make our text more accessible and uniform}, and also we omit the verification of equality (i.e. the judgementally-equal syntax needs to be mapped to the same data for well-defined-ness).

Note that, since every piece is intrinsic well-typed, $\denotesS{t}$ means the same as $\denotesS{\goodTerm{\Gamma}{t}{T}}$ (i.e. every term is tracking its context and type)

@@

% We omit the model here for extensible inductive type and inheritance judgement because they require inductive facility in meta-logic. 
We spell-out the model for extensible type and inheritance judgement here, and both of them require inductive facility in meta-logic. In set theory, that would be using \textit{Rule sets}~\citep{timany2017consistency,aczel1998relating}, but we are using Agda's Inductive facility instead. 
% In appendix we will use the native inductive notion from Agda to construct models. 
As expected, we use an inductive type with the constructor from W-type to model our extensible inductive type, and the recursor for the inductive type is directly modelling the recursor for our extensible inductive type. For inheritance judgement, we inductively define an almost-identical syntactic structure as model, (but not a QIIT), and we inductively define meta-level substitution to implement the explicit substitution. We also define $inh$ operation by induction strictly following $\beta$-law. Basically both ideas are using the syntax of themselves in meta-logic to model, because its definition is pretty "well-founded". Most importantly, this doesn't affect our aim to model bottom type using empty set to derive consistency.

Notice that we interpret the bottom type using empty set, and thus we know it is not possible to derive $\cdot \vdash t : \bot$: otherwise we have $\cdot \models_S t : \bot$ which exactly means $\denotesS{t}(\star) \in \denotesS{\bot}(\star) = \emptyset$ 
and that is a contradiction.

### 2022-06-05 syntax.tex
Save.

\judgebox{\goodWSig{\Gamma}{\sigma}{n}}
$$
\Rule[name=Inductive Type]
{\goodWSig{\Gamma}{\sigma}{n}}
{\goodType{\Gamma}{W\sigma}{i}}
\quad
\Rule[name=Inductive Sig]
{\goodWSig{\Delta}{\sigma}{n}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodWSig{\Gamma}{\sigma[\gamma]}{n}
  \quad \goodType{\Gamma}{W (\sigma[\gamma]) \equiv (W \sigma)[\gamma]}{i}}
$$

@@

Finally we introduce the inheritance judgement. Inheritance is represented by judgement so naturally first order data (i.e. not something function can return), but it itself looks very much alike a function that transform a given linkage.

>>>>

\judgebox{\goodWSig{\Gamma}{\sigma}{n}}
$$
\Rule[name=Emp WSig]
{}
{\goodWSig{\Gamma}{w\cdot}{0}}
\quad
\Rule[name=WSig Add]
{\goodWSig{\Gamma}{s}{n}
  \quad \goodType{\Gamma}{A}{i}
  \quad \goodType{\Gamma, A}{B}{i}}
{\goodWSig{\Gamma}{w^+ \  s \  A \  B}{n+1}}
\quad
\Rule[name=Ind Univ]
{\goodWSig{\Gamma}{\sigma}{n}}
{\goodType{\Gamma}{\bW \sigma}{i}}
$$

$$
\Rule[name=WSig Proj]
{\goodWSig{\Gamma}{s}{n} \quad j < n}
{\goodType{\Gamma}{\pi^j_1 s}{i} \quad \goodType{\Gamma, \pi^j_1 s}{\pi^j_2  s}{i}}
\quad
\Rule[name=Ind Sig]
{\goodWSig{\Delta}{\sigma}{n}
  \quad {\goodSub{\Gamma}{\gamma}{\Delta}}}
{\goodWSig{\Gamma}{\sigma[\gamma]}{n}
  \quad \goodType{\Gamma}{W (\sigma[\gamma]) \equiv (W \sigma)[\gamma]}{i}}
$$

$$
\Rule[name=Ind Type]
{\goodTerm{\Gamma}{T}{\bW \sigma}}
{\goodTerm{\Gamma}{W \ T}{\cU}}
\quad
\Rule[name=Ind Term]
{\goodTerm{\Gamma}{T}{\bW \sigma}
  \quad \goodTerm{\Gamma}{a}{\pi^j_1\sigma}
  \quad \goodTerm{\Gamma, \pi^j_1\sigma}{b}{\pi^j_2\sigma}}
{\goodTerm{\Gamma}{Wsup \ T \ a \ b}{El\ (W\ T)} }
$$

@@

Finally we introduce the inheritance judgement. Inheritance are judgement so naturally second class citizen data (i.e. not something function can return), and it itself looks very much alike a function that transform a given linkage.

The syntax of the inheritance judgement is very close to how Family is defined in the surface syntax -- a Family with no parent is an inheritance with empty input; an extended Family is exactly an inheritance -- because the parent of an extended Family can be overridden thus inheritance should only define upon the ``interface'' instead of a concrete family. However, since our current formulation doesn't support further-binding (i.e. we cannot do inheritance inside a family), our formulation still have some distance to the real family polymorphism.
